# Week 6: Analysis of longitudinal data

## Reading the datasets, factoring and structures
```{r, message = FALSE}

library(ggplot2); library(dplyr); library(tidyr)

RATSL <- read.csv(file = "data/RATSL.csv")
RATSL$ID <- factor(RATSL$ID)
RATSL$Group <- factor(RATSL$Group)
str(RATSL)

BPRSL <- read.csv(file = "data/BPRSL.csv")
BPRSL$treatment <- factor(BPRSL$treatment)
BPRSL$subject <- factor(BPRSL$subject)
str(BPRSL)

```

In RATS data we have three groups, who were put on different diets. Then each rat was weighted repeatedly
over a nine week period. The question of most interest is whether the growth profiles of the
three groups differ. 

In BPRS data we had two different treatment groups, whose BPRS (the psychiatric rating scale) was measured on 9 weeks. The question of interest is whether two different treatments differed.


## 1) Analyses on RATS data

### Graphs
``` {r}

# Graphical display of RATS

ggplot(RATSL, aes(x = Time, y = weight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATSL$weight), max(RATSL$weight))) +
  ggtitle("Individual response profiles by group for the RATSL data")


```
  
We see that in all groups the weigth seems to increase when time increases. In groups 2 and 3 increase is more clear. Graph shows us that there seems to be a difference in baseline weights at least between Group 1 vs other Groups. Groups 2 ja 3 are more similar. In every group there is one unsual individual, who we migth delete later.

### Standardized graphs
``` {r}
RATSL <- RATSL %>%
  group_by(Time) %>%
  mutate(stweigth = (weight - mean(weight))/sd(weight) ) %>%
  ungroup()

glimpse(RATSL)

ggplot(RATSL, aes(x = Time, y = stweigth, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:16, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(name = "standardized weigth") +
  ggtitle("Individual response profiles by group for RATSL data after standardization")

```
  
With standardized values we see more clearly, that there is one outlier observartion in each group. 

In Group 1 the weigth profile seem to be pretty much the same over time points. In Group 3 there is a small decrease in weigths over time. Only in Group 2 the weigth profiles seem to increase.

### Summary graph
``` {r}

# Number of time, baseline (week 0) included
n <- RATSL$Time %>% unique() %>% length()

# Summary data with mean and standard error of bprs by treatment and week 
RATSS <- RATSL %>%
  group_by(Group, Time) %>%
  summarise(mean = mean(weight), se = sd(weight)/sqrt(n) ) %>%
  ungroup()

# Glimpse the data
glimpse(RATSS)

# Plot the mean profiles
ggplot(RATSS, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.4) +
  theme(legend.position = c(0.8,0.4)) +
  scale_y_continuous(name = "mean(weigth) +/- se(weigth)") +
  ggtitle("Mean response profiles for the three groups in the RATS data.")


```
  
The mean profiles plot shows that the error bars do not overlap.

### Mean as a summary measure
``` {r}

# Create a summary data by treatment and subject with mean as the summary variable (ignoring baseline week 0).
RATSL10S <- RATSL %>%
  filter(Time > 0) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(weight) ) %>%
  ungroup()

# Glimpse the data
glimpse(RATSL10S)

# Draw a boxplot of the weigth mean versus group
ggplot(RATSL10S, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(weigth), Times 8-64")

```
  
Here is a plot of mean as a summary variable.  
We can see three outlier observations, one in each group. Let's remove these and draw the plot again.

```{r}
# Create a new data by filtering the outlier and adjust the ggplot code the draw the plot again with the new data
RATSL10S1 <- RATSL10S %>%
  filter((mean < 550 & mean > 500 )|  (mean > 250 & mean < 475))


glimpse(RATSL10S1)

ggplot(RATSL10S1, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(weigth), Times 8-64")


```
 
Now without those three observations, the group summaries look very good.  

Next we can test the group growth difference with t-test.

### Two sample t-test, linear model and anova
``` {r}

# Perform a two-sample t-test
t.test(mean ~ Group == 1  | Group ==2, data = RATSL10S1, var.equal = TRUE)
t.test(mean ~ Group == 1  | Group ==3, data = RATSL10S1, var.equal = TRUE)
t.test(mean ~ Group == 2  | Group ==3, data = RATSL10S1, var.equal = TRUE)


# Add the baseline from the original data as a new variable to the summary data
RATS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt"        , sep = "\t", header = TRUE)
               
RATSL10S2 <- RATSL10S %>%
  mutate(baseline = RATS$WD1)

# Fit the linear model with the mean as the response 
fit <- lm(mean ~ baseline + Group, data = RATSL10S2)

# Compute the analysis of variance table for the fitted model with anova()
anov <- anova(fit)
anov

```
From t-tests we see that between Groups 1 and 2 and 2 and 3 the mean in weights differ. Between group 1 and 3 there is no significant difference.  

This means that Group 2 diet is the one that caused an significant growth in weigth.

From Anova we see that the baseline RATS is strongly related to the RATS values taken after treatment has begun, but there is no evidence of a group difference even after conditioning on the baseline value.


## 2) Analyses on BPRS data

### Plotting
```{r}

# plotting again 
ggplot(BPRSL, aes(x = week, y = bprs, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)))


```
  
From the plot we see, that between the two treatments there seem to no clear difference. With both bprs measure decreases with time, so treatments seem to work.

### Linear regression model
```{r}
# let's fit a linear regression
BPRS_reg <- lm(bprs ~ week + treatment, data = BPRSL)

# print out a summary of the model
summary(BPRS_reg)
BPRS_reg

```
Treatment2 doesn't seem to differ from Treatment1, because p-value is 0.66.  
The the p-value of week is very small, so week (time) has an negative effect on bprs. This means both treatments seems to work, but not differently from eachother.  
However this model assumes independence of the repeated measures of bprs, and this assumption is highly unlikely, so let's move to other models.

### Random intercept model
```{r}

# access library lme4
library(lme4)

# Create a random intercept model
BPRS_ref <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRSL, REML = FALSE)

# Print the summary of the model
summary(BPRS_ref)

```

### Random Intercept and Random Slope Model
```{r}
# create a random intercept and random slope model
BPRS_ref1 <- lmer(bprs ~ week + treatment + (week | subject), data = BPRSL, REML = FALSE)

# print a summary of the model
summary(BPRS_ref1)



# perform an ANOVA test on the two models
anova(BPRS_ref1, BPRS_ref)

```
P-value is 0.02636, which is pretty low (< 0.05). So we can say that the random intercept and slope model
provides a better fit for these data.

### A random intercept and random slope model with the interaction
```{r}
# create a random intercept and random slope model with the interaction
BPRS_ref2 <- lmer(bprs ~ week * treatment + (week | subject), data = BPRSL, REML = FALSE)

# print a summary of the model
summary(BPRS_ref2)

# perform an ANOVA test on the two models
anova(BPRS_ref2, BPRS_ref1)

```
P-value is 0.07495 (> 0.05), which means the random intercept and slope model provides a better fit than the interaction model for these data.

### Plotting again
```{r}

# draw the plot of BPRSL with the observed bprs values
ggplot(BPRSL, aes(x = week, y = bprs, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)))

```


```{r}
# Create a vector of the fitted values
Fitted <- fitted(BPRS_ref1)

# Create a new column fitted to RATSL

BPRSL <- mutate(BPRSL, fitted = Fitted)

# draw the plot of BPRSlL with the Fitted values of weight

ggplot(BPRSL, aes(x = week, y = fitted, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRSL$fitted), max(BPRSL$fitted)))


```
  
This graphic underlines how well the random intercept and slope model fits the observed data. Since the model is linear, it looks pretty different from the observed data where there is more curves. However, as a simplification it seems to work okay.

